Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-07 rmd/manuscript.Rmd
update: mansucript

Under each scenario of omnivory, we crossed disturbance (frequency and intensity), productivity, and dispersal. Disturbance frequency and intensity were controlled via $p_m$ ( `r unique(df_sim$p_disturb) %>% sort()`) and $\mu_m$ (`r unique(df_sim$mean_disturb_source) %>% sort()`), respectively. Spatial variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(df_sim$sd_disturb_source)`), and local disturbance variation ($\sigma_{l}=$ `r min(df_sim$sd_disturb_lon)`). This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries while allowing strong within-tributary downstream correlation, a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. We used the basal species' growth rate $r_b$ (`r unique(df_sim$r_b)`) as a proxy for resource availability as it resembles the gross primary production. For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(df_sim$theta)`; long-distance dispersal $\theta=$ `r min(df_sim$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(df_sim$p_dispersal)`).

@@

There were two consistent patterns in the relationship between FCL and ecosystem geometry (size and complexity). First, the effects of ecosystem size and complexity were modulated by disturbance and omnivory. Second, resource availability and dispersal had limited influences on the qualitative relationship between FCL and ecosystem geometry, although they had important direct effects on the FCL. Below we describe these results in detail.

## Interactive effects of ecosystem geometry, disturbance, and omnivory

@@

The exception is the "chain" scenario (i.e., no omnivory) with no disturbance. In this scenario, ecosystem size had a clear positive effect on FCL (**Figure 2**) while ecosystem complexity had a negative influence (**Figure 3**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increases the total carrying capacity whereas ecosystem complexity decreases it (**Figure XXX**). However, these patterns characteristic to the "chain" scenario diminished when considering some levels of omnivory because FCLs with omnivory were less reliant on the carrying capacity of the basal species.

State transition shed lights on how FCL changes occurred. In the chain motif, the "additive" mechanism (i.e., the addition of higher trophic levels) is always responsible for the changes in FCL. With disturbance, state transition occurred from "B" to "B + C" as ecosystem size or complexity increases. With no disturbance, state transition was different between size and complexity. Ecosystem size facilitated the transition from "B + C" to "B + C + P", while ecosystem complexity drove the opposite transition ("B + C + P" $\rightarrow$ "B + C").

In the weak and strong omnivory motifs, ecosystem size and complexity increased FCL through different mechanisms. Ecosystem size increased FCL primarily through the additive mechanism, facilitating the state transition from "B" to "B + P", particularly with frequent disturbance. However, the "insertion" mechanism (the transition from "B + P" to "B + C + P") also took place in the strong omnivory motif with rare disturbance ($p_m = 0.05$). In contrast, the insertion mechanism was responsible for the effect of ecosystem complexity on FCL. In the weak omnivory scenario, the transition from "B + P" to "B + C + P" was predominant. In the strong omnivory scenario, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change. However, as disturbance frequency increased, the primary process switched to the additive mechanism ("B" $\rightarrow$ "B + P").

## Productivity and dispersal

Productivity and dispersal had limited influences on the qualitative relationship between FCL and ecosystem geometry. However, they had important direct effects on FCL. Productivity influences were dependent on omnivory. Overall, greater productivity increased (chain & strong omnivory) or decreased FCL (weak omnivory), although the effect size varied with ecosystem geometry and disturbance levels. Dispersal distance had a positive effect on FCL, particularly in scenarios with frequent and intense disturbance.

xxx

@@

The most significant finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain scenario). However, as the frequency of intense disturbance increases, small and/or simple networks could not maintain long food chains because habitats were too homogeneous with limited disturbance refugia. While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

The synergy between ecosystem geometry and disturbance has also an implication for why the dynamical stability hypothesis [@pimmNumberTrophicLevels1977] has been least supported by the literature. While increased disturbance frequency consistently reduced FCL, its magnitude depended strongly on ecosystem geometry. This observation should provide a possible explanation for the limited support for the dynamical stability hypothesis, because such an interaction has been rarely assessed in previous studies. Although FCL research often assumes that the effects of environmental controls are independent of one another, exploring potential interactions among them could provide useful insights into what controls FCL in nature.

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, when the chain motif was combined with no-disturbance, ecosystem size and complexity had contrasting influences on FCL. It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, the total carrying capacity within a network dictates FCL. This effect of ecosystem size is understandable because it increases the total carrying capacity. In contrast, ecosystem complexity, here assessed as branching probability, decreased the total carrying capacity as more branched networks have disproportionate numbers of headwater habitats with small local carrying capacities. Thus, ecosystem complexity had an opposite effect on FCL with the chain motif. However, the dependence of FCL on local carrying capacity diminished in weak and strong omnivory scenarios. This fact explains the lack of positive (ecosystem size) or negative FCL-geomtery relationship (ecosystem complexity) in these motifs with no disturbance.

Importantly, however, disturbance overwhelmed the effect of omnivory on the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result provides a possible explanation for why the ecosystem size hypothesis has been consistently supported in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that, if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. However, such substantial food web reorganization may have limited influences on how ecosystem geometry dictates FCL.

The effect of productivity was context-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif, this effect was vague (strong omnivory) or even reversed (weak omnivory) in omnivory motifs. With omnivory, high productivity effectively increases the abundance of the basal species, stimulating the prey switching of the intraguild predator from the intraguild prey to the basal species. As such, increased productivity does not necessarily increase FCL. Our results suggest that this prediction from non-spatial models can be applied to spatially complex ecosystems.

>>>>

Under each scenario of omnivory, we crossed disturbance (frequency and intensity), productivity, and dispersal. Disturbance frequency and intensity were controlled via $p_m$ ( `r unique(df_sim$p_disturb) %>% sort()`) and $\mu_m$ (`r unique(df_sim$mean_disturb_source) %>% sort()`), respectively. Spatial variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(df_sim$sd_disturb_source)`), and local disturbance variation ($\sigma_{l}=$ `r min(df_sim$sd_disturb_lon)`). This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries while allowing strong within-tributary downstream correlation, a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. We used a combination of basal species' growth rate $r_b$ and base carrying capacity $K_{base}$ (low productivity: $r_b =$ `r min(df_sim$r_b)`, $K_{base} =$ `r min(df_sim$base_k)` ; high productivity = $r_b =$ `r max(df_sim$r_b)`, $K_{base} =$ `r max(df_sim$base_k)` ) as a proxy for resource availability. For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(df_sim$theta)`; long-distance dispersal $\theta=$ `r min(df_sim$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(df_sim$p_dispersal)`).

@@

There were two consistent patterns in the relationship between FCL and ecosystem geometry (size and complexity). First, the effects of ecosystem size and complexity were modulated by disturbance. Second, productivity strengthened the relationship between FCL and ecosystem geometry with important direct effects that were motif dependent. Third, dispersal had a minimal but non-negligible positive effect on FCL. Below we describe these results in detail.

## Interactive effect of ecosystem geometry and disturbance

@@

The exception is the "chain" scenario (i.e., no omnivory) with no disturbance and low productivity. In this scenario, ecosystem size had a positive effect on FCL (**Figure 2**) while ecosystem complexity had a negative influence (**Figure 3**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increased the total carrying capacity whereas ecosystem complexity decreased it (**Figure XXX**). However, these patterns characteristic to the "chain" scenario diminished when considering some levels of omnivory.

State transitions shed lights on how FCL changes occurred. In the chain motif, the "additive" mechanism (i.e., the addition of higher trophic levels) was always responsible for the changes in FCL. With disturbance, state transition occurred from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases. Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

In the weak and strong omnivory motifs, ecosystem size and complexity increased FCL through different mechanisms. Ecosystem size increased FCL primarily through the additive mechanism, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (Figure XXX). In contrast, the insertion and additive mechanisms were responsible for the effect of ecosystem complexity on FCL, depending on simulation scenarios. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (Figure XXX). However, the additive mechanism ("No species" $\rightarrow$ "B + P") seems to be dominant in the high productivity scenario (Figure XXX). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (Figure XXX). As disturbance frequency increased, the primary process switched to the additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; Figure XXX).

## Productivity effect

High productivity strengthened the relationship between FCL and ecosystem geometry. In the chain motif, productivity greatly increased the overall FCL (Figure XXX), thereby shifting the upper bound of FCL (**Figures 2 and 3**). In contrast, increased productivity greatly decreased the minimum levels of FCL with weak and strong omnivory with little influence on the maximum FCL (**Figures 2, 3, and XXX**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 2 and 3**). Such patterns were particularly clear with long-distance dispersal (**Figures 2 and 3**).

State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" (i.e., additive) was dominant regrardless of disturbance frequency. The introduction of omnivory allowed the transition between "B + P" and "B + C + P", the primary direction of the transition differed between weak and strong omnivory. With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL. However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of "No species" state (Figure XXX), especially in small and/or simple ecosystems (Figure XXX).

## Dispersal effect

Dispersal distance had a positive influence on FCL, regardless of simulation scenarios. However, this direct effect was rather weak in comparison to other factors.

State transitions occurred through either additive (chain) or insertion mechanisms (omnivory). In the chain motif, the transitions were either "B" $\rightarrow$ "B + C" or "B + C" $\rightarrow$ "B + C+ P", depending on disturbance frequency. In the omnivory motifs, the predominant transition was from "B + P" to "B + C + P."

@@

The most significant finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain motif with low productivity). However, as the frequency of intense disturbance increases, small and/or simple networks became unable to maintain long food chains. Here, increased ecosystem size and complexity may give greater resistance to disturbance through different mechanisms. Ecosystem size simply increases the "total amount" of disturbance refugia, from which dispersing propagules recolonize impacted habitat patches. In the meantime, ecosystem complexity may increases the connectivity to disturbance refugia. Indeed, branching shortens the mean distance between patches and therefore may enhance immigration from less undisturbed patches. It is also conceivable that the proportion of disturbance refugia increased in more branched networks. However, we did not observe a clear proportional increase in disturbance refugia with increasing branching probability. Therefore, each of geometric features support long food chains through distinct mechanisms. While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

The synergy between ecosystem geometry and disturbance has also an important implication for why the dynamical stability hypothesis [@pimmNumberTrophicLevels1977] has been least supported by the literature. Although increased disturbance frequency consistently reduced FCL, its magnitude depended strongly on ecosystem geometry. This observation should provide a possible explanation for the limited support for the dynamical stability hypothesis in field studies. Although FCL research often assumes that the effects of environmental controls are independent of one another, exploring potential interactions among them could provide useful insights into what controls FCL in nature.

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, with no disturbance and low productivity, FCL of the chain motif had contrasting relationships with ecosystem size (positive) and complexity (negative). It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, FCL followed the changes in the total carrying capacity along the gradients of ecosystem size and complexity. However, these relationships diminished in omnivory motifs as the dependence of FCL on local carrying capacity decreased (**Figure XXX**). Importantly, however, disturbance overwhelmed the effect of omnivory on the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result provides a possible explanation for why the ecosystem size hypothesis has been consistently supported in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that, if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. However, such substantial food web reorganization may have limited influences on how ecosystem geometry dictates FCL.

The effect of productivity was context-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif, this effect was vague (strong omnivory) or even reversed (weak omnivory) in omnivory motifs. Intraguild predation plays an important role in explaining this context dependent effect. In the high productivity scenario with omnivory, the increased abundance of the basal species enhances the intraguild predator. This may subsequently facilitate the local exclusion of the intraguild prey through increased predation pressure, if the attack rate is sufficiently high [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. Indeed, in the weak omnivory motif (the high value of the attack rate of P on C), we observed the state transition from "B + C + P" to "B + P" that shortened FCL. However, in the strong omnivory motif, the transition was
